An open and continuously updated fern tree of life (FTOL)

### 2022-02-15 ms/manuscript.Rmd
Update plots
Add clade codes to FTOL plot

# Load packages
library(conflicted)
library(rmarkdown)
library(bookdown)
library(targets)
library(assertr)
library(ape)
library(scales)
library(tidyverse)
library(rgnparser)
library(ggtree)
library(glue)
library(Polychrome)
library(patchwork)
library(magrittr)
library(ggrepel)
library(ggokabeito)

@@

# Make tibble of backbone tips with higher-level taxonomy
backbone_taxonomy <-
	tibble(species = get_tips_in_ape_plot_order(plastome_tree_rooted)) %>%
	# Add higher taxonomy
	left_join(sanger_sampling, by = "species") %>%
	# Add column for phylogenetic order
	mutate(phy_order = seq_len(nrow(.))) %>%
	# Outgroup lacks family data, so call "og"
	mutate(
		family = case_when(
			outgroup == TRUE ~ "og",
			TRUE ~ family),
		major_clade = case_when(
			outgroup == TRUE ~ "og",
			TRUE ~ major_clade)
	)
```

@@

# Format coverage at order-ish level for plotting
b <-
	coverage_data %>%
	filter(rank == "major_clade") %>%
	left_join(major_clade_exemp_taxa, by = c(name = "major_clade")) %>%
	arrange(phy_order) %>%
	rename(clade = name) %>%
	mutate(
		clade = fct_reorder(clade, phy_order) %>%
			fct_rev(),
		coverage_text = glue("{number(n, accuracy = 1)}/{number(n_accepted, accuracy = 1)}")
	) %>%
	ggplot(aes(x = coverage, y = clade, fill = clade)) +
	geom_col(color = "grey10", size = 0.2) +
	geom_text(
		aes(label = coverage_text),
		hjust = 0,
		nudge_x = 0.01,
		size = 6*.36 # 6 pt font
	) +
	scale_x_continuous(
		labels = percent,
		limits = c(0,1),
		expand = expansion(mult = c(0, 0))
	) +
	scale_fill_manual(values = clade_cols) +
	labs(x = "Coverage") +
	theme_bw() +
	theme(
		axis.text.y = element_text(size = 6),
		axis.title.y = element_blank(),
		legend.position = "none",
		plot.background = element_rect(fill = "transparent", colour = NA),
		panel.grid.major.y = element_blank(),
		panel.grid.minor.x = element_blank()
	)

# Tree plot ----
phy <- plastid_tree_dated

# Make tibble for annotating major clades
clade_tibble <-
	tibble(
		Polypodiineae = c("Leucostegia_truncata", "Didymochlaena_fijiensis"),
		Aspleniineae = c("Acystopteris_taiwaniana", "Rhachidosorus_pulcher"),
		Dennstaedtiineae = c("Monachosorum_flagellare", "Dennstaedtia_auriculata"),
		Pteridineae = c("Coniogramme_affinis", "Pteris_dispar"),
		Lindsaeineae = c("Cystodium_sorbifolium", "Odontosoria_retusa"),
		Saccolomatineae = c("Saccoloma_inaequale", "Orthiopteris_campylura"),
		Cyatheales = c("Cyathea_bipinnatifida", "Culcita_macrocarpa"),
		Salviniales = c("Marsilea_nashii", "Salvinia_oblongifolia"),
		Schizaeales = c("Schizaea_elegans", "Lygodium_palmatum"),
		Gleicheniales = c("Matonia_foxworthyi", "Sticherus_bifidus"),
		Hymenophyllales = c("Cephalomanes_javanicum", "Hymenophyllum_tortuosum"),
		Marattiales = c("Angiopteris_evecta", "Danaea_erecta"),
		`Ophioglossales + Psilotales` = c("Psilotum_nudum", "Sceptridium_nipponicum"),
		Osmundales = c("Osmundastrum_cinnamomeum", "Todea_barbara"),
		Equisetales = c("Equisetum_bogotense", "Equisetum_hyemale")
	) %>%
	pivot_longer(names_to = "major_clade", values_to = "taxa", cols = everything()) %>%
	arrange(major_clade) %>%
	group_by(major_clade) %>%
	summarize(taxa = list(taxa)) %>%
	mutate(
		mrca = map_chr(taxa, ~getMRCA(phy, .))
	) %>%
	transmute(
		major_clade,
		mrca,
		dummy = ""
	) %>%
	mutate(mrca = parse_number(mrca))

a <- ggtree(
	phy, 
	layout = "circular", right = FALSE, size = 0.1,
) +
	geom_cladelab(
		data = clade_tibble,
		mapping = aes(
			node = mrca,
			color = major_clade,
			label = dummy
		),
		offset = 0.1,
		barsize = 2) +
	scale_color_manual(values = clade_cols) +
	theme(legend.position = "none")

ftol_plot <-
	a + inset_element(b, left = 0.85, bottom = 0.14, right = 1.3, top = .8) +
	plot_annotation(
		theme = theme(
			plot.margin = unit(c(-2,5,-2,-1.6), "cm"))
	)
```

@@

**`r figure("ftol")`**. Fern Tree of Life (FTOL). Inset plot shows coverage by major clade (order or suborder). Numbers next to each bar show sampled species out of total number of species. Taxonomy follows [@PteridophytePhylogenyGroupI2016].

@@

```{r div-times-make}
# Convert family ages to long format
family_ages_median <-
family_stem_ages %>%
	select(family, current_best = age) %>%
	left_join(other_dates, by = "family") %>%
	mutate(family = fct_reorder(family, current_best)) %>%
	pivot_longer(values_to = "age", -family) %>%
	separate(name, into = c("author", "age_type"), sep = "_") %>%
	assert(not_na, author, age_type) %>%
	filter(age_type %in% c("best", "median"))

@@

div_times_plot <-
ggplot(mapping = aes(y = family)) +
	geom_errorbarh(
		data = family_ages_errors,
		aes(xmax = high, xmin = low, color = author),
		position = position_dodge(width = 0.5)
	) +
	geom_point(
		data = filter(family_ages_median, age_type %in% c("best", "median")),
		aes(x = age, color = author, shape = author),
		position = position_dodge(width = 0.5)) +
	scale_shape_manual(
		values = ages_shapes,
		limits = ages_sampling$source, 
		labels = ages_sampling$label) +
	scale_color_manual(
		values = ages_colors,
		limits = ages_sampling$source, 
		labels = ages_sampling$label) +
	labs(x = "Age (millions of years)") +
	guides(shape = guide_legend(nrow = 2)) +
	theme_bw() +
	theme(
		axis.title.y = element_blank(),
		legend.title = element_blank(),
		legend.position = "bottom"
	)
```

@@

**`r figure("divtimes")`**. Stem age of fern families (mya) estimated by selected studies.
For studies that used methods with confidence intervals, error bars indicate lower and upper 95% highest posterior density levels and point indicates median [@Rothfels2015a; @Testo2016a].
For other studies, point indicates best (most likely) estimate [@Schuettpelz2007b\; this study].

>>>>

# Load packages
library(conflicted)
library(rmarkdown)
library(bookdown)
library(targets)
library(assertr)
library(ape)
library(scales)
library(tidyverse)
library(rgnparser)
library(ggtree)
library(glue)
library(Polychrome)
library(patchwork)
library(magrittr)
library(deeptime)
library(ggrepel)
library(ggokabeito)
library(ggtext)

@@

# Make tibble of backbone tips with higher-level taxonomy
backbone_taxonomy <-
	tibble(species = get_tips_in_ape_plot_order(plastome_tree_rooted)) %>%
	# Add higher taxonomy
	left_join(sanger_sampling, by = "species") %>%
	# Add column for phylogenetic order
	mutate(phy_order = seq_len(nrow(.))) %>%
	# Outgroup lacks family data, so call "og"
	mutate(
		family = case_when(
			outgroup == TRUE ~ "og",
			TRUE ~ family),
		major_clade = case_when(
			outgroup == TRUE ~ "og",
			TRUE ~ major_clade)
	)

clade_cols_tib <-
	tibble(
		color = clade_cols,
		major_clade = names(clade_cols)
	) %>%
	mutate(
		code = substr(major_clade, 1, 2),
		code = case_when(
			major_clade == "Saccolomatineae" ~ "Sac",
			major_clade == "Salviniales" ~ "Sal",
			TRUE ~ code
		)) %>%
	assert(is_uniq, code)

```

@@

# Format y-axis text: show major clade abbreviations in bold
clade_labs_tib <-
	clade_cols_tib %>%
	select(major_clade, code) %>%
	mutate(
		clade_short = map2_chr(major_clade, code, ~str_remove(.x, .y)),
		label = glue("<b>{code}</b>{clade_short}"))

clade_labs <- clade_labs_tib$label %>%
	set_names(., clade_labs_tib$major_clade)

# Format coverage at order-ish level for plotting
b <-
	coverage_data %>%
	filter(rank == "major_clade") %>%
	left_join(major_clade_exemp_taxa, by = c(name = "major_clade")) %>%
	arrange(phy_order) %>%
	rename(clade = name) %>%
	mutate(
		clade = fct_reorder(clade, phy_order) %>%
			fct_rev(),
		coverage_text = glue("{number(n, accuracy = 1)}/{number(n_accepted, accuracy = 1)}")
	) %>%
	ggplot(aes(x = coverage, y = clade, fill = clade)) +
	geom_col(color = "grey10", size = 0.2) +
	geom_text(
		aes(label = coverage_text),
		hjust = 0,
		nudge_x = 0.01,
		size = 6*.36 # 6 pt font
	) +
	scale_x_continuous(
		labels = percent,
		limits = c(0,1),
		expand = expansion(mult = c(0, 0))
	) +
	scale_fill_manual(values = clade_cols) +
	scale_y_discrete(labels = clade_labs) +
	labs(x = "Coverage") +
	theme_bw() +
	theme(
		axis.text.y = element_markdown(size = 6),
		axis.title.y = element_blank(),
		legend.position = "none",
		plot.background = element_rect(fill = "transparent", colour = NA),
		panel.grid.major.y = element_blank(),
		panel.grid.minor.x = element_blank()
	)

# Tree plot ----
# Make tibble for annotating major clades
clade_tibble <- 
	# Use assess_monophy to get MRCA of each major clade
	# They should all by monophyletic anyways
	select(sanger_sampling, species, major_clade) %>%
	assess_monophy(
		taxon_sampling = .,
		tree = plastid_tree_dated,
		tax_levels = "major_clade"
	) %>%
	get_result_monophy(1) %>%
	filter(!taxon %in% c(
		"Selaginellales", "Lycopodiales", "Isoëtales", "Isoetales")) %>%
	verify(all(monophyly == "Yes")) %>%
	transmute(major_clade = taxon, mrca = parse_number(mrca)) %>%
	left_join(clade_cols_tib, by = "major_clade")

# Helper function to label clades with different offset
# For some reason, using the same offset.text around the plot
# results in different positions of text relative to bars. 
# Account for this by manual adjusting offset.text
ftol_geom_cladelab <- function(code_select, offset_text) {
	geom_cladelab(
		data = filter(clade_tibble, code %in% code_select),
		mapping = aes(
			node = mrca,
			color = major_clade,
			label = code
		),
		offset = 0.1,
		offset.text = offset_text,
		fontsize = 2.5,
		textcolour = "black",
		barsize = 2)
}

a <- ggtree(
	plastid_tree_dated, layout = "circular", right = FALSE, size = 0.1) + 
	ftol_geom_cladelab(
		clade_tibble$code[clade_tibble$code %in% c("Eq", "Op", "Ma", "Os", "Gl")], 10) +
	ftol_geom_cladelab(
		clade_tibble$code[clade_tibble$code %in% c("Hy", "Sc", "Sal")], 14) +
	ftol_geom_cladelab(
		clade_tibble$code[clade_tibble$code %in% c("Cy", "Sac", "Li", "De")], 17) +
	ftol_geom_cladelab("Pt", 26) +
	ftol_geom_cladelab("As", 34) +
	ftol_geom_cladelab("Po", 17) +
	scale_color_manual(values = clade_cols) +
	theme(legend.position = "none")

# Assemble plots ----
ftol_plot <-
	a + inset_element(b, left = 0.85, bottom = 0.14, right = 1.3, top = .8) +
	plot_annotation(
		theme = theme(
			plot.margin = unit(c(-2,5,-2,-1.4), "cm"))
	)
```

@@

**`r figure("ftol")`**.
Fern Tree of Life (FTOL).
Tree rooted on bryophytes.
Inset plot shows coverage by major clade (order or suborder).
Bold part of each clade name is its code, which is also indicated on the tree.
Numbers next to each bar show sampled species out of total number of species.
Taxonomy follows [@PteridophytePhylogenyGroupI2016].

@@

```{r div-times-make}
# Convert family ages to long format
family_ages_median <-
family_stem_ages %>%
	select(family, current_best = age) %>%
	left_join(other_dates, by = "family") %>%
		left_join(
		unique(select(sanger_sampling, family, major_clade)),
		by = "family"
	) %>%
	left_join(clade_cols_tib, by = "major_clade"
	) %>%
	mutate(family = fct_reorder(family, current_best)) %>%
	pivot_longer(values_to = "age", -c(family, color, code, major_clade)) %>%
	separate(name, into = c("author", "age_type"), sep = "_") %>%
	assert(not_na, author, age_type) %>%
	filter(age_type %in% c("best", "median")) %>%
	assert(not_na, major_clade, color)

@@

# Format y-axis text: show major clade in parentheses
family_labs_tbl <-
family_ages_median %>%
	select(family, code) %>%
	unique() %>%
	mutate(label = glue("{family} ({code})"))
family_labs <- family_labs_tbl$label %>%
	set_names(., family_labs_tbl$family)

# Provide names of periods in caption
# Make sure to read and check that they are all there!
period_names <-
	periods %>%
	# Filter to range including Ordovician to Neogene
	filter(min_age < 475) %>%
	filter(min_age > 1) %>%
	arrange(desc(min_age)) %>%
	mutate(label = glue("{abbr} ({name})")) %>%
	pull(label) %>%
	paste(collapse = ", ")

div_times_plot <-
ggplot(mapping = aes(y = family)) +
	geom_errorbarh(
		data = family_ages_errors,
		aes(xmax = high, xmin = low, color = author),
		position = position_dodge(width = 0.5)
	) +
	geom_point(
		data = filter(family_ages_median, age_type %in% c("best", "median")),
		aes(x = age, color = author, shape = author),
		position = position_dodge(width = 0.5)) +
	scale_shape_manual(
		values = ages_shapes,
		limits = ages_sampling$source, 
		labels = ages_sampling$label) +
	scale_color_manual(
		values = ages_colors,
		limits = ages_sampling$source, 
		labels = ages_sampling$label) +
	scale_y_discrete(labels = family_labs) +
	scale_x_reverse("Age (millions of years)") +
	coord_geo(
		dat = list("periods", "eras"),
		pos = list("b", "b"),
		abbrv = list(TRUE, FALSE),
		xlim = c(500, 0),
		fill = "transparent",
		color = "grey20",
		lab_color = "grey30",
		height = unit(1, "lines"),
		size = 3) +
	guides(shape = guide_legend(nrow = 2)) +
	theme_bw() +
	theme(
		axis.title.y = element_blank(),
		legend.title = element_blank(),
		legend.position = "bottom"
	)
```

@@

**`r figure("divtimes")`**.
Stem age of fern families (mya) estimated by selected studies.
For studies that used methods with confidence intervals, error bars indicate lower and upper 95% highest posterior density levels and point indicates median [@Rothfels2015a; @Testo2016a].
For other studies, point indicates best (most likely) estimate [@Schuettpelz2007b\; this study].
Codes in parentheses after family names indicate major clade as in `r figure("ftol")`.
Period name abbreviations as follows: `r period_names`.
